---
title: Analyze breast cancer single-cell data from Basses et al. (Nature Medicine 2021)
author: "M. Andreatta <massimo.andreatta at unil.ch> and S. Carmona <santiago.carmona at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

This dataset contains single-cell data from pre- versus on-treatment biopsies from patients receiving anti-PD1 immunotherapy (n=29) or chemotherapy (n=11). TCR-seq data are also available. Lacking an objective measure of response to therapy, in this study they use as a proxy the level of T cell expansion to infer response to treatment.

Paper: <https://www.nature.com/articles/s41591-021-01323-8> Dataset: <https://biokey.lambrechtslab.org/en/data-access> Internal link: <https://drive.switch.ch/index.php/s/uTrKHnTMohuUori>

```{r, message=FALSE, warning=F, results=FALSE}
renv::restore()
#remotes::install_github("carmonalab/ProjecTILs")

library(ProjecTILs)
library(scGate)
library(Seurat)

library(gridExtra)
library(ggplot2)
library(reshape2)
```

Read in data

Tcell matrices have T cell subtype annotation by the authors

```{r}
ddir <- "input/Bassez_breast_data"

#Count matrices
f1 <- sprintf("%s/1864-counts_tcell_cohort1.rds", ddir)

cohort1 <- readRDS(f1)
dim(cohort1)

#Metadata
meta1 <- read.csv(sprintf("%s/1870-BIOKEY_metaData_tcells_cohort1_web.csv", ddir))
rownames(meta1) <- meta1$Cell

head(meta1)

data.c1 <- CreateSeuratObject(cohort1, project="Cohort1_IT", meta.data = meta1)

```

TCR metadata

```{r}
# #TCR metadata # tcr1 <- read.csv(sprintf("%s/1879-BIOKEY_barcodes_vdj_combined_cohort1.csv", ddir))  # tcr1$barcode <- paste0(tcr1$barcode, "-1") # rownames(tcr1) <- tcr1$barcode #  # data.c1 <- AddMetaData(data.c1, metadata = tcr1) #  # table(data.c1$SAMPLE_ID, data.c1$patient_id) # #Note: some patients (e.g. BIOKEY_13 do not have TCR_seq data) #  # #Calculate clomnotype counts and frequencies #  # clonotypes <- table(data.c1$clonotype) # clonotypes.vs.sample <-  as.matrix(table(data.c1$clonotype, data.c1$SAMPLE_ID)) # clonotypes.sums <- apply(clonotypes.vs.sample, 2, sum) #  # freqs <- vector() #  # for (i in 1:length(clonotypes)) { #    ctype <- names(clonotypes)[i] #    count <- clonotypes[i] #    sample <- gsub("(\\S+)_clonotype\\S+", "\\1", ctype, perl=T) #    freqs[ctype] <- count/clonotypes.sums[sample] # } #  # ctypes.vector <- data.c1@meta.data[,c("clonotype")] # freqs.order <- rep(NA,length(ctypes.vector)) # counts.order <- rep(NA,length(ctypes.vector)) #  # for (r in 1:length(ctypes.vector)) { #    c <- ctypes.vector[r] #    if (!is.na(c)) { #       counts.order[r] <- clonotypes[c] #       freqs.order[r] <- freqs[c] #    } # } # data.c1$clone_count <- counts.order # data.c1$clone_freq <- freqs.order #  # hist(data.c1$clone_count) # hist(data.c1$clone_freq) #  # #Define levels of expansion # data.c1$expand <- NA # data.c1$expand[!is.na(data.c1$clone_count) & data.c1$clone_count==1] <- "Single" # data.c1$expand[!is.na(data.c1$clone_count) & data.c1$clone_count>1 & data.c1$clone_count<=5] <- "Small(1-5)" # data.c1$expand[!is.na(data.c1$clone_count) & data.c1$clone_count>5 & data.c1$clone_count<=10] <- "Medium(6-10)" # data.c1$expand[!is.na(data.c1$clone_count) & data.c1$clone_count>11 & data.c1$clone_count<=20] <- "Large(11-20)" # data.c1$expand[!is.na(data.c1$clone_count) & data.c1$clone_count>20] <- "Hyper(>20)" #  # data.c1$expand <- factor(data.c1$expand,  #                 levels = c("Hyper(>20)", "Large(11-20)", "Medium(6-10)", "Small(1-5)", "Single", NA)) #  # table(data.c1$expand, useNA = "ifany")`
```

# QC

Ribosomal and mitochondrial content

```{r fig.height=3}
percent.ribo.dv <- PercentageFeatureSet(data.c1, pattern = "^RP[LS]")
percent.mito.dv <- PercentageFeatureSet(data.c1, pattern = "^MT-")

data.c1 <- AddMetaData(data.c1, metadata = percent.ribo.dv, col.name = "percent.ribo")
data.c1 <- AddMetaData(data.c1, metadata = percent.mito.dv, col.name = "percent.mito")

Idents(data.c1) <- "patient_id"
VlnPlot(data.c1, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito"), ncol = 2, pt.size=0)

Idents(data.c1) <- "cellSubType"
VlnPlot(data.c1, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito"), ncol = 2, pt.size=0)

quantile(data.c1$nFeature_RNA, c(0, 0.01, 0.1, 0.5, 0.9, 0.99))
quantile(data.c1$nCount_RNA, c(0, 0.01, 0.1, 0.5, 0.9, 0.99))
```

Minimal QC to remove extreme outliers

```{r}
dim(data.c1)
data.c1 <- subset(data.c1, subset = nFeature_RNA>500 & nFeature_RNA<6000 & 
                       nCount_RNA>600 & nCount_RNA< 25000 &
                       percent.ribo < 50  &
                       percent.mito < 15)
dim(data.c1)
```

Have a look at the data, unsupervised analysis

```{r}
#First a little downsampling to speed things up...
#ds <- 20000
#data.c1 <- subset(data.c1, cells = sample(Cells(data.c1), ds))
library(SignatuR)


data.c1 <- NormalizeData(data.c1, verbose = FALSE)
data.c1 <- FindVariableFeatures(data.c1, selection.method = "vst", nfeatures = 1000, verbose = FALSE)

cycleCycle.symbol <- c(SignatuR$Hs$Programs$cellCycle.G1S, SignatuR$Hs$Programs$cellCycle.G2M)
                       
mito.genes <- grep(pattern = "^MT-", rownames(data.c1), value = TRUE)
ribo.genes <- grep(pattern = "M?RP[LS]|DAP3|FAU|UBA52", rownames(data.c1), value = TRUE)

genes.remove <- c(cycleCycle.symbol, mito.genes, ribo.genes) 

data.c1@assays$RNA@var.features <- setdiff(data.c1@assays$RNA@var.features, genes.remove)
```

```{r}
set.seed(1234)
data.c1 <- ScaleData(data.c1)
data.c1 <- RunPCA(data.c1, features = data.c1@assays$RNA@var.features, ndims.print = 1:5, nfeatures.print = 5)

data.c1 <- RunUMAP(data.c1, reduction = "pca", dims = 1:20, seed.use=123)
DimPlot(data.c1, reduction = "umap", group.by = "patient_id") + ggtitle("Patient")
DimPlot(data.c1, reduction = "umap", group.by = "cellSubType", label=T) + ggtitle("Cell type")
DimPlot(data.c1, reduction = "umap", group.by = "timepoint") + ggtitle("Timepoint")
DimPlot(data.c1, reduction = "umap", group.by = "expansion") + ggtitle("Expansion")
DimPlot(data.c1, reduction = "umap", group.by = "BC_type") + ggtitle("Cancer type")
```

#Check some markers on UMAP

```{r fig.height=3}
FeaturePlot(data.c1, reduction = "umap", features=c("CD4","CD8B","FOXP3","CD3E"), pt.size = 0.01, slot = "data", ncol=2)
FeaturePlot(data.c1, reduction = "umap", features=c("PDCD1","TCF7","GZMB","GZMK"), pt.size = 0.01, slot = "data", ncol=2)
FeaturePlot(data.c1, reduction = "umap", features=c("CXCL13","CD40LG","BCL6","CD200"), pt.size = 0.01, slot = "data", ncol=2)
FeaturePlot(data.c1, reduction = "umap", features=c("HAVCR2","ENTPD1","PRF1","IL7R"), pt.size = 0.01, slot = "data", ncol=2)
FeaturePlot(data.c1, reduction = "umap", features=c("TRDC","TRGC1","FGFBP2","KLRF1"), pt.size = 0.01, slot = "data", ncol=2)
FeaturePlot(data.c1, reduction = "umap", features=c("MKI67","TOP2A","IFNG","nFeature_RNA"), pt.size = 0.01, slot = "data", ncol=2)
```

The CD4_EX cluster appears to be quite heterogeneous. It expresses Pdcd1, but also a good portion does express Tcf7, some Tim3+/Prf1+ cells in the north. Also specific expression of Cd200, which may indicate that (some of) these cells are Tfh...?

Remove NK and gd T cells

```{r}
table(data.c1$cellSubType)

dim(data.c1)
data.ab <- subset(data.c1, subset=cellSubType %in% c("NK_REST","Vg9Vd2_gdT","gdT","NK_CYTO"), invert=T)
dim(data.ab)
```

#ProjecTILs

```{r}
path <- "~/Dropbox/CSI/Datasets/projectTils/"
ref.cd8 <- readRDS(sprintf("%s/NickB.Human_CD8_TILs_atlas.rds",path))

cols <- ref.cd8@misc$atlas.palette

# Plotting
ref.umap <- DimPlot(ref.cd8,  group.by = 'functional.cluster', label = T, repel = T, cols = cols) + theme(aspect.ratio = 1)
ref.umap
```

```{r}
query.list <- SplitObject(data.ab, split.by="patient_id")
query.projected <- Run.ProjecTILs(ref.cd8, query = query.list, ncores=8)
```

If projecting by patient, merge back results after ProjecTILs analysis

```{r}
query.merged <- suppressMessages(Reduce(ProjecTILs:::merge.Seurat.embeddings, query.projected))
```

Save projected object

```{r}
saveRDS(query.merged, "aux/Bassez.projected.merged.rds")
query.merged <- readRDS("aux/Bassez.projected.merged.rds")
```

```{r}
g <- DimPlot(query.merged, group.by = "cellSubType", label=T) +
  theme(aspect.ratio=1) + theme_bw()
```

Compare CD8T subtypes

```{r fig.height=3}
query.merged <- subset(query.merged, subset=cellSubType %in% c("CD8_EM","CD8_EMRA","CD8_EX","CD8_EX_Proliferating","CD8_N","CD8_RM"))
library(patchwork)
query.bysub <- SplitObject(query.merged, split.by = "cellSubType")
pll <- list()

for (n in names(query.bysub)) {
  x <- query.bysub[[n]]
  pll[[n]] <- plot.projection(ref.cd8, x, linesize = 0.5, pointsize = 0) + ggtitle(n) +
    theme(axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(), aspect.ratio = 1)
}
wrap_plots(pll)

ggsave("plots/Bassez_projected_by_subtype.png", height=6, width=12)
```

```{r fig.height=4, fig.width=8}
genes4radar <- c('CD4','CD8A', 'LEF1', "TCF7", "CCR7", "GZMK", "FGFBP2",'FCGR3A','KLRG1',"PDCD1",
                                  'ZNF683','ITGAE', "CRTAM", "CD200","HAVCR2",
                                  "TOX", "ENTPD1", 'SLC4A10','TRAV1-2')

plot.states.radar(ref.cd8, query = query.bysub, min.cells = 20,
                  genes4radar = genes4radar) 

```

```{r fig.height=4, fig.width=8}
#By subtype
for (n in names(query.bysub)) {
  pll <- plot.states.radar(ref.cd8, query = query.bysub[[n]], min.cells = 30,
                  genes4radar = genes4radar, return.as.list = TRUE)
  
  p <- wrap_plots(pll) + plot_annotation(sprintf("Profiles for %s", n))
  plot(p)
  
  ggsave(sprintf("plots/Bassez_radars_by_subtype.%s.pdf", n), height=9, width=14)
}
```

Compare responders vs. non-responders (based on clonal expansion) - pre-therapy

```{r fig.height=3.5}
#Subset on pre-treatment samples
query.sub <- subset(query.merged, subset=timepoint=="Pre")
#Split by expansion (defined by authors)
query.sub <- subset(query.sub, subset=expansion %in% c("E","NE"))
query.sub.byExp <- SplitObject(query.sub, split.by = "expansion")

a <- plot.projection(ref.cd8, query=query.sub.byExp$E, linesize = 0.3, pointsize = 0) + 
  ggtitle("Expanded") + NoLegend() +  theme(aspect.ratio = 1,
        axis.text=element_blank(),
        axis.ticks=element_blank())
b <- plot.projection(ref.cd8, query=query.sub.byExp$NE, linesize = 0.3, pointsize = 0)  +
  ggtitle("Non-Expanded") + NoLegend() + theme(aspect.ratio = 1,
        axis.text=element_blank(),
        axis.ticks=element_blank())

a | b

ggsave("plots/Bassez_expanded_vs_non.png", height=4.5, width=6)

c <- plot.statepred.composition(ref.cd8, query.sub.byExp$E, metric = "percent") + ylim(0,30) + ggtitle("Expanded")
d <- plot.statepred.composition(ref.cd8, query.sub.byExp$NE, metric = "percent") + ylim(0,30)  + ggtitle("Non-Expanded")
c | d
```

Fold-change

```{r fig.heigh=8, fig.width=5, fig.align="center",  cache=FALSE}
t <- table(query.merged$functional.cluster)>20
which.types <- names(t[t==TRUE])

cols_use  <- ref.cd8@misc$atlas.palette[which.types]

#Responder vs non Responder
query.list <- SplitObject(query.merged, split.by = "expansion")

norm.c <- table(query.list[["NE"]]$functional.cluster)/sum(table(query.list[["NE"]]$functional.cluster))
norm.q <- table(query.list[["E"]]$functional.cluster)/sum(table(query.list[["E"]]$functional.cluster))

foldchange <- norm.q[which.types]/norm.c[which.types]
foldchange <- sort(foldchange,decreasing = T)

tb.m <- melt(foldchange)
colnames(tb.m) <- c("Cell_state","Fold_change")
ggplot(tb.m, aes(x=Cell_state, y=Fold_change, fill=Cell_state)) + geom_bar(stat="identity") +
  scale_fill_manual(values=cols_use) + geom_hline(yintercept = 1) + scale_y_continuous(trans='log2') +
  theme(axis.text.x=element_blank(), legend.position="left") + ggtitle("Responder vs. Non-responder") +
  theme_light()

ggsave("plots/Bassez_expanded_vs_non.foldchange.png", height=4, width=4)

```

Compare responders vs. non-responders (based on clonal expansion) - post-therapy

```{r fig.height=3.5}
#Subset on pre-treatment samples
query.sub <- subset(query.merged, subset=timepoint=="On")
#Split by expansion (defined by authors)
query.sub <- subset(query.sub, subset=expansion %in% c("E","NE"))
query.sub.byExp <- SplitObject(query.sub, split.by = "expansion")

a <- plot.projection(ref.cd8, query=query.sub.byExp$E, linesize = 0.5, pointsize = 0.2) + ggtitle("Expanded")
b <- plot.projection(ref.cd8, query=query.sub.byExp$NE, linesize = 0.5, pointsize = 0.2)  + ggtitle("Non-Expanded")

a | b

c <- plot.statepred.composition(ref.cd8, query.sub.byExp$E, metric = "percent") + ylim(0,40) + ggtitle("Expanded")
d <- plot.statepred.composition(ref.cd8, query.sub.byExp$NE, metric = "percent") + ylim(0,40)  + ggtitle("Non-Expanded")
c | d
```

Plot clonal expansion on the UMAP

```{r}
palette <- colorRampPalette(c("#FF4B20", "#FFB433", "#C6FDEC", "#7AC5FF", "#0348A6"))
DimPlot(query.merged, group.by = "expansion") + scale_color_manual(values = c(palette(5)), na.value="grey")
```

Show largest clones

```{r fig.width=12, fig.height=12}
clone_call = "clonotype"   #which column to use to determine clonotypes
top.clones <- 20

clonotypeSizes <- sort(table(query.merged[[clone_call]]),decreasing = T)

bigClonotypes <- names(clonotypeSizes)[1:top.clones]

plots <- list()
for (i in 1:length(bigClonotypes)) {
  
  ctype <- bigClonotypes[i]
  patient <- unique(query.merged$patient_id[which(query.merged[[clone_call]]==ctype)])
  aa <- unique(query.merged$cdr3_aa[which(query.merged[[clone_call]]==ctype)])
  
  plots[[i]] <- plot.projection(ref.cd8, query.merged[,which(query.merged[[clone_call]]==ctype)],
                                linesize=0.3, pointsize = 0.3) + 
    ggtitle(sprintf("%s\n%s\n%s - size %i", ctype, aa, patient, clonotypeSizes[ctype]))
}

wrap_plots(plots)

```

Export representative plots

```{r fig.width=6}
a <- plots[[1]] + NoLegend() + theme(axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(), aspect.ratio = 1,
        plot.title = element_text(size=8))

b <- plots[[7]] + NoLegend() + theme(axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(), aspect.ratio = 1,
        plot.title = element_text(size=8))

c <- plots[[14]] + NoLegend() + theme(axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(), aspect.ratio = 1,
        plot.title = element_text(size=8))

a | b | c

#ggsave("plots/Bassez_largest_clonotypes.png", height=3.5, width=11)
ggsave("plots/Bassez_selected_clonotypes.png", height=3.5, width=11)
```

```{r fig.height=4}
cc <- c("BIOKEY_28_Pre_clonotype2","BIOKEY_28_Pre_clonotype1")

cc <- "BIOKEY_14_Pre_clonotype1"
#cc <- "BIOKEY_11_On_clonotype1"

sub <- subset(query.merged, subset=clonotype %in% cc)
plot.projection(ref.cd8, sub)

genes4radar <- c('LEF1', "TCF7", "CCR7", "IL7R",'KLRG1',
                 'ZNF683',"TOX", 'ITGAE',"GZMK","CXCR3","LY6C2", "GZMB","PRF1", "PDCD1", "HAVCR2",
                 "ENTPD1","LAG3","CTLA4")


plot.states.radar(ref.cd8, sub, genes4radar = c(genes4radar,"ITGA1","IL7R","GZMB"), min.cells = 10)
```
